Fixing Vulnerabilities Hinders Maintainability

### 2019-02-03 paper/ICPC19/icpc19.tex
393 and grammarly pass

\begin{abstract}
Security is a requirement of utmost importance to produce software with high
quality. However, there is still a considerable amount of zero-day
vulnerabilities being discovered, and fixed, almost weekly. We suspect that when
developers address these vulnerabilities on their codebases, they actually
affect the maintainability of software, potentially introducing other
vulnerabilities. This paper evaluates the impact of refactorings to improve
security on the maintainability of open-source software. Maintainability is
measured based on the \emph{Better Code Hub}'s model of 10 guidelines using a
dataset including 607 security commits. Results show that fixing software
vulnerabilities hinders the maintainability of open-source software. This is
particularly the case for refactorings that address \emph{Broken Authentication
and Session Management} issues, \emph{Memory Leaks} and \emph{Denial-of-Service}
attacks. Furthermore, we conclude that changes to codebases to fix vulnerabilities
need to be performed with extra care.
\end{abstract}

@@

Software quality is important because it is ultimately related to the overall
cost for developing, extending and maintaing software applications. Software
quality characteristics include, but are not limited to, functional correctness,
reliability, usability, maintainability, and security. Security is an important
non-functional requirement during the development of software systems.  However,
there is still a considerable amount of zero-day vulnerabilities being
discovered, and fixed, almost weekly as disclosed by the Zero Day Initiative
website\footnote{Zero Day Initiative website available  at
https://www.zerodayinitiative.com/advisories/published/ (Accessed on \today{})}.
These vulnerabilities are gateways for attackers to penetrate
systems and exploit their resources.
As an example, in the last week of January 2019, a severe vulnerability in Apple
FaceTime was revealed\footnote{https://thehackernews.com/2019/01/apple-facetime-privacy-hack.html
(Accessed on \today{})} --- the outcome of a design/logical flaw that allowed callers to see and
hear others without them picking up the call.

In 2011, the International Organization for Standardization (ISO) issued an
update for software product quality ISO/IEC 25010~\cite{iso:2011} considering
\emph{Security} as one of the main software product quality characteristics.
Unmaintainable code is hard to test, analyse and re-use, and thus more difficult
to find vulnerabilities and defects. Therefore, writing maintainable code
prevents the introduction of vulnerabilities in the codebases. In this work, we are
concerned about the impact that patching vulnerabilities may have in the
open-source software maintainability. Are developers affecting software
maintainability when trying to improve software safety? We suspect that some of
the patches may have a negative impact in the software maintainability and
possibly even be the cause of the introduction of new vulnerabilities.

ISO describes software maintainability as ``the degree of effectiveness and
efficiency with which a software product or system can be modified to improve
it, correct it or adapt it to changes in environment, and in requirements'' on
software quality ISO/IEC 25010. As ISO does not provide any specific guidelines or
formula to calculate maintainability, we resort on Software Improvement
Group (SIG)'s web-based source code analysis service Better Code Hub
(BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given
project. There are other well-known standards and models that have been proposed to increase software safety:
Common Criteria~\cite{common:2009} which suffered a lot of criticism regarding the
costs associated and poor technical evaluation; the OWASP
Application Security Verification Criteria~\cite{oswap:2009} which is focused only
on web applications, and a model proposed by Xu et al. (2013)~\cite{6616351} for
raiting software security which we believe it was one of the first steps taken
by SIG to introduce security on their maintainability model.

BCH compiles the ISO/IEC 25010 into a set of 10 guidelines that are the
base of BCH's evaluation. Two examples are the
\emph{Write Simple Units of Code} based on the McCabe Complexity~\cite{1702388}
and \emph{Keep your Codebase Small} based on the idea that smaller codebases are
easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.
% Compliance to the guidelines is derived from the SIG's
% industry benchmark which, according to official sources, consists of over 8
% billion lines of code in more than 180 different technologies. SIG analyzes
% around 15 million lines of code, per week.

The particular subject of this paper is to explore whether there is a trade-off
between refactoring source code to patch disclosed vulnerabilities while keeping
(or even improving) the maintainability of software systems. Hedegus et al. (2018)~\cite{HEGEDUS2018313},
analyzed the difference of maintainability between
groups of refactored elements and non-refactored and conclude that the source
code elements subjected to refactorings had significantly lower maintainability
than the elements not affected.

In this paper, we present the results of our analysis on the maintainability of
607 security changes collected from open-source software, from different
programming languages. The main contributions of the present work is:
\begin{itemize}
	\item An empirical study on the impact of security changes on software
	maintainability and what patterns need more attention.

@@

As an example, consider the refactoring of the Datagram Transport Layer Security
(DTLS) protocol implementation in OpenSSL\footnote{OpenSSL is a tookit that
contains open-source implementations of the SSL and TLS cryptographic
protocols.} to address a vulnerability where an attacker could cause a Denial-of-Service
(DoS) attack by crafting DTLS handshake messages to trigger memory allocations
corresponding to large length values. This vulnerability is listed at the Common
Vulnerabilities and Exposures dictionary as CVE-2014-3506\footnote{CVE-2014-3506
details available at http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506
(Accessed on \today{})}. It is amongst the vulnerabilities studied in our
evaluation. The following snippet presents the changes performend on the
\emph{ssl/d1\_both.c} file\footnote{CVE-2014-3506 fix available  at
https://github.com/openssl/openssl/commit/\\1250f12613b61758675848f6600ebd914ccd7636
(Accessed on \today{})} by the OpenSLL developers to patch the vulnerability.

@@

\setcounter{lstannotation}{0}
\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
\\CVE-2014-3506 vulnerability},label={lst:vuln}]
+ static unsigned long dtls1_max_handshake_message_len(const SSL *s){ /*!\annotation{lst:func1}!*/
+  unsigned long max_len = DTLS1_HM_HEADER_LENGTH + SSL3_RT_MAX_ENCRYPTED_LENGTH;
+  if(max_len < (unsigned long)s->max_cert_list)
+     return s->max_cert_list;
+  return max_len;
+ }

@@

Every SSL/TLS connection begins with a handshake which is responsible by the negotiation between two parties.
DTLS provides a mechanism for fragmenting a handshake message over a number of packets, each of which can be
transmitted separately. In the example (Listing~\ref{lst:vuln}), all functions receive as input the
SSL session, \texttt{s}. In the \texttt{dtls1\_reassemble\_fragment} function the size of the fragment \texttt{msg\_hdr-$>$frag\_off+frag\_len} is checked against the handshake message size \texttt{msg\_hdr->msg\_len}
(line 13) and then against the maximum handshake message size allowed by DTLS1 for \texttt{s} (line 16-21).
However, it is not verified if \texttt{msg\_hdr->msg\_len} is lower than the allowed by DTLS. Thus, in line 37,
when \texttt{msg\_len} is allocated to the fragment buffer, more memory than the allowed might be allocated. According to the message commit, $10$ handshake messages are allowed per DTLS connection. Hence, a remote attacker could leverage this issue to trigger
memory allocations corresponding to large length values and lead the system to memory exhaustion that could end
in a DoS attack. The code changes performed to fix the vulnerability were the following:

~\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was inserted to return the number of
maximum number of bytes allowed in a DTLS handshake message for \texttt{s}. The mininum is $16$KB, but may be greater
if the maximum certificate list size requires it which is set by the user. More specifically, it checks if the
maximum certificate list size
(\texttt{s-$>$max\_cert\_list}) requires more length than the maximum length
(\texttt{max\_len}). If that condition holds, the function returns the size of
the certificate list, otherwise it returns the maximum length. This function replaces lines $18$-$21$.

~\ref{lst:func5} and~\ref{lst:func4} add the check to different functions to ensure that the message
and the fragment length are never higher than the maximum allowed by DTSL. However, \ref{lst:func5} is
the responsible for patching the vulnerability.

@@

Often security flaws require refactoring code to make software more secure.
However, there is no evidence yet of how security refactorings impact the
maintainability of open-source software. Our suspicion is that developers tend
to introduce technical debt in their software when refactoring the codebose to
address a security flaw because they tend to choose the easiest path to solve
it. To address it, we compute the maintainability of $607$ commits using
\emph{Better Code Hub}. The same approach is applied to a randomly generated
dataset of regular commits --- baseline --- that we use to understand how
maintainability evolves when security refactorings are performed versus when
they are not.

@@

\end{lstlisting}
%
\section{Methodology}\label{sec:methodology}
%
The methodology used to measure the impact of security refactorings on the
maintainability of open-source software is discussed this section. The methodology
comprehends the following steps, as illustrated in Figure~\ref{fig:met}:
\begin{enumerate}
	\item Extract relevant data from a dataset containing
$716$ security refactorings collected from open-source software available on
GitHub~\cite{Reis:2017:IJSSE}.
	\item A baseline of regular commits was
randomly collected from the list of projects of the main dataset to evaluate the impact of regular commits on the maintainability of open-source
software.
\item Use the Software Improvement Group's\footnote{SIG's website is available
at https://www.sig.eu/ (Accessed on \today{})} web-based source code analysis
service \emph{Better Code Hub} (BCH)\footnote{BCH's website is available at
https://bettercodehub.com/ (Accessed on \today{})} to quantify
maintainability for both security and regular commits.
\end{enumerate}

%
\begin{figure}[h]
 	\centering 	\includegraphics[width=0.5\textwidth]{figures/methodology.pdf}
 	\caption{Study Methodology}
	\label{fig:met}
\end{figure}
%
\subsection{Dataset}
%
To evaluate the impact of security refactorings on the maintainability of
open-source software, we use a dataset of security flaws which is the outcome of
mining $94$ GitHub projects. Reis and Abreu (2017) mined open-source
software aiming at the extraction of real examples --- created by real
developers --- to test and assess the performance of static analysis tools since
using hand-seeded test cases or mutations could lead to misleading assessments
of the capabilities of the tools~\cite{just2014mutants}. The study yielded a
dataset with $716$ test cases for $16$ security patterns. Some of the patterns
are based in the OWASP Top 10 of 2013~\cite{oswap:2013} and OWASP Top 10 of
2017~\cite{oswap:2017}. Each test case of the
dataset is a triplet: the commit before the refactoring, the commit responsible
for the refactoring, and the snippets of code that differ from one version to
another (typically, called \textit{diffs}) --- where one can easily review the
code used to fix the security flaw. In this study, we focus on computing the
maintainability of the commits before and after the security refactoring to
evaluate if the impact was positive, negative or none.

@@

The $716$ refactorings in the dataset were run against the BCH toolset to
calculate their maintainability reports. However, there are refactorings
discarded from tour study due to limitations of BCH: in particular, lack of
language support and project size. The final dataset used in this paper contains
$607$ security refactorings, labelled as one of the following categories (we
have decided to change the label to \textit{Miscellaneous} of the patterns in
the dataset containing less than $20$ instances):

\begin{itemize}
	\item \textbf{Using Components with Known Vulnerabilities.} The majority of
	software produced today integrates several components, such as libraries and
	frameworks, which may affect the software if developers use vulnerable
	versions - usually known and disclosed somewhere on the Internet.
	\textit{Based on the OWASP Top 10 of 2017.}

	\item \textbf{Broken Authentication \& Session Management.} Uncorrectly
	implemented functionalities related to authentication and session management,
	allowing attacker to gain access to session tokens, passwords, keys and other
	sensitive data. \textit{Based on the OWASP Top 10 of 2013.}

	\item \textbf{Cross-Site Request Forgery.} Poor session tokens generation and
	management usually allow attackers to send forged HTTP requests including
	authentication information from the victim to vulnerable web application.
	\textit{Based on the OWASP Top 10 of 2013.}

	\item \textbf{Denial-of-Service} Security flaws that can allow a attacker to
	flood or crashing services. This attacks usually occur when the system
	receives much traffic causing it to slow down and eventually stop. For
	example, flaws allowing attackers to trigger memory allocations corresponding
	to large length values (Listing~\ref{lst:vuln}).

	\item \textbf{Miscellaneous} This pattern comprises several others categories
	of security refactorings (e.g., \textit{path traversal and buffer overflow}).
	It contains several other security refactorings that do not have their own
	pattern yet. The patterns not satisfying the size requirement of more than 20
	refactorings are contemplated here: \textit{Resource Leak},
	\textit{SHA-1 Hash Function}, \textit{Broken Access Control}, \textit{Sensitive Data Exposure},
	\textit{Insufficient Attack Protection}, \textit{Overflow}, \textit{Path Traversal}.

	\item \textbf{Cross-Site Scripting.} Lack of proper validation or escaping
	allow attackers to submit untrusted data to web browsers through malicious
	scripts that can hijack the user sessions or redirect the user to malicious
	sites. \textit{Based on the OWASP Top 10 of 2017.}

	\item \textbf{Injection.} When developers do not keep untrusted data separate
	from commands and queries. If a attacker sends a string that exploits the
	syntax of the interpreter, then an injection attack is possible (e.g., SQL and
	LDAP injection). \textit{Based on the OWASP Top 10 of 2017.}

@@

The dataset contemplates security flaws for more than 13 different languages
being PHP (39\%) and C (24\%) the most prevalent ones; see Figure~\ref{fig:lang}.
To classify the applications in the dataset, we use a taxonomy for open-source
software from previous work~\cite{7816479}. Figure~\ref{fig:domain}
presents the projects domain distribution: \textit{Aplication Software} ($25$),
software that provides end-users with functional systems; \textit{Web libraries
and frameworks} ($20$); \textit{System Software} ($18$), software that provides
services and infrustructures (e.g., operating systems, servers and databases);
\textit{Software Tool} ($13$), software that supports development (e.g.,
programming languages, compilers, package managers, IDEs); and, \textit{Non-web
libraries and frameworks} ($18$), for desktop and mobile software.
%
Table~\ref{tab:dataset} presents the descriptive statistics of the $94$
open-source projects involved in this study, including the number of forks, the number of
stars, the number of watchers, the number of contributors, the number of commits, the number of
branches, the number of releases,  size (cf. GitHub it is the size of the whole
repository, including all of its history, in kilobytes), the number of issues, and the
number of pull requests.

%
\subsection{Security vs. Baseline Commits}
%
Previous studies attempted to measure the impact of regular refactorings on
open-source software maintainability~\cite{HEGEDUS2018313}. However,
there is no previous work focused on comparing the impact of security
refactorings with regular refactorings on maintainability.
We analyze the maintainability of regular commits
(i.e., commits not related with security fixes) and use them as a baseline.

The baseline dataset uses the security commits dataset as input. For each
security commit, one random commit is selected from the list of all commits of
that particular project. We originate the regular commits from the security
commits to ensure that differences in maintainability are not consequence of
characteristics of different projects.
%
\subsection{Maintainability Analysis}

As said before, the web-based source code analysis service \emph{Better Code
Hub} (BCH) is used to collect the maintanability reports of the refactorings of
each project. Table~\ref{tab:guidelines} presents the $10$ guidelines proposed
by BCH's authors for delivering software that is not difficult to
maintain~\cite{Visser:2016:OREILLY}. During each guideline evaluation, the tool
determines the compliance towards one guideline by establishing limits for the
percentage of code allowed to be in each of the $4$ risk severity levels
(\emph{low risk}, \emph{medium risk}, \emph{high risk} and \emph{very high
risk}). If the project does not violate the thresholds, then it is compliant
with the guideline. These thresholds are calibrated by BCH using their own
data/experience --- open-source and closed software systems. If a project is
compliant with a guideline, it means that it is at least $65\%$ better than the
software used by BCH to calculate the thresholds\footnote{Check the answer to
\emph{How can I adjust the threshold for passing/not passing a guideline?} at
https://bettercodehub.com/docs/faq (Accessed on \today{})}.

Figure~\ref{fig:bchrep} shows an example of the report
provided by BCH for a project after finishing its evaluation. The examples
refers to the OpenSSL CVE-2014-3506 vulnerability refactoring ---
as described as motivating example in Section~\ref{sec:motivation}. This
version of OpenSSL only complies with 1 out of 10 guidelines: \emph{Write
Clean Code}.

@@

\toprule
\textbf{10 Guidelines} & \textbf{Description}\\
\midrule
\textbf{Write Short Units of Code} & Limit code units to 15 LOCs because smaller
 units are easier to understand, reuse and test them\\\midrule
\textbf{Write Simple Units of Code} & Limit branch points to 4 per unit because
it makes units easier to test and modify \\\midrule
\textbf{Write Code Once} & Do not copy code because bugs tend to replicate at
multiple places (inefficient and error-prone)\\\midrule
\textbf{Keep Unit Interfaces Small} & Limit the number of parameters to at most
4 because it makes units easier to understand and reuse\\\midrule
\textbf{Separate Concerns in Modules} & Avoid large modules because changes in
loosely coupled databases are easier to oversee and execute\\\midrule
\textbf{Couple Architecture Components Loosely} & Minimize the amount of code
within modules that is exposed to modules in other components\\\midrule
\textbf{Keep Architecture Components Balanced} & Balancing the number of
components ease locating code and allow for isolated maintenance\\\midrule
\textbf{Keep your Codebase Small} & Reduce and avoid the system size because
small products are easier to manage and maintain\\\midrule
\textbf{Automate Tests} & Test your codebase because it makes development
predictable and less risky\\\midrule
\textbf{Write Clean Code} & Avoid producing software with code smells because
it is more likely to be maintainable in the future\\
\bottomrule
\end{tabular}
\label{tab:guidelines}
\end{table}

\begin{figure}[h]
 	\centering 	\includegraphics[width=0.5\textwidth]{figures/bch_report.png}
 	\caption{Maintainability report of OpenSSL's CVE-2014-3506 vulnerability
refactoring for the guideline \emph{Write Simple Units of Code} provided by
\emph{Better Code Hub}. This version of OpenSSL does not comply with the
guideline in the example since the bars do not reach the threshold points. This
example only complies with $\sfrac{1}{10}$ guidelines (\emph{Write Clean Code}).}
	\label{fig:bchrep}
\end{figure}

>>>>

\begin{abstract}
Security is a requirement of utmost importance to produce software with high
quality. However, there is still a considerable amount of zero-day
vulnerabilities being discovered, and fixed, almost weekly. We suspect that when
developers address these vulnerabilities on their codebases, they actually
affect the maintainability of software, potentially introducing other
vulnerabilities. This paper evaluates the impact of refactorings to improve
security on the maintainability of open-source software. Maintainability is
measured based on the \emph{Better Code Hub}'s model of $10$ guidelines using a
dataset including $607$ security commits. Results show that fixing software
vulnerabilities hinders the maintainability of open-source software. This is
particularly the case for refactorings that address \emph{Broken Authentication
and Session Management} issues, \emph{Memory Leaks} and \emph{Denial-of-Service}
attacks. Furthermore, we conclude that changes to codebases to fix vulnerabilities
need to be performed with extra care.
\end{abstract}

@@

Software quality is important because it is ultimately related to the overall
cost of developing, extending and maintaining software applications. Software
quality characteristics include, but are not limited to, functional correctness,
reliability, usability, maintainability, and security. Security is an important
non-functional requirement during the development of software systems.  However,
there is still a considerable amount of zero-day vulnerabilities being
discovered, and fixed, almost weekly as disclosed by the Zero Day Initiative
website\footnote{Zero Day Initiative website available at
https://www.zerodayinitiative.com/advisories/published/ (Accessed on \today{})}.
These vulnerabilities are gateways for attackers to penetrate
systems and exploit their resources.
As an example, in the last week of January $2019$, a severe vulnerability in Apple
FaceTime was revealed\footnote{https://thehackernews.com/2019/01/apple-facetime-privacy-hack.html
(Accessed on \today{})} --- the outcome of a design/logical flaw that allowed callers to see and
hear others without them picking up the call.

In $2011$, the International Organization for Standardization (ISO) issued an
update for software product quality ISO/IEC 25010~\cite{iso:2011} considering
\emph{Security} as one of the main software product quality characteristics.
Unmaintainable code is hard to test, analyze and re-use, and thus more difficult
to find vulnerabilities and defects. Therefore, writing maintainable code
prevents the introduction of vulnerabilities in the codebases. In this work, we are
concerned about the impact that patching vulnerabilities may have in the
open-source software maintainability. Are developers affecting software
maintainability when trying to improve software safety? We suspect that some of
the patches may have a negative impact on the software maintainability and
possibly even be the cause of the introduction of new vulnerabilities.

ISO describes software maintainability as ``the degree of effectiveness and
efficiency with which a software product or system can be modified to improve
it, correct it or adapt it to changes in environment, and in requirements'' on
software quality ISO/IEC 25010. As ISO does not provide any specific guidelines or
formula to calculate maintainability, we resort on Software Improvement
Group (SIG)'s web-based source code analysis service Better Code Hub
(BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given
project. There are other well-known standards and models that have been proposed to increase software safety:
Common Criteria~\cite{common:2009} which suffered a lot of criticism regarding the
costs associated and poor technical evaluation; the OWASP
Application Security Verification Criteria~\cite{oswap:2009} which is focused only
on web applications, and a model proposed by Xu et al. ($2013$)~\cite{6616351} for
rating software security which we believe it was one of the first steps taken
by SIG to introduce security on their maintainability model.

BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
base of BCH's evaluation. Two examples are the
\emph{Write Simple Units of Code} based on the McCabe Complexity~\cite{1702388}
and \emph{Keep your Codebase Small} based on the idea that smaller codebases are
easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.

The particular subject of this paper is to explore whether there is a trade-off
between refactoring source code to patch disclosed vulnerabilities while keeping
(or even improving) the maintainability of software systems. Hedegus et al. ($2018$)~\cite{HEGEDUS2018313},
analyzed the difference of maintainability between
groups of refactored elements and non-refactored and conclude that the source
code elements subjected to refactorings had significantly lower maintainability
than the elements not affected.

In this paper, we present the results of our analysis on the maintainability of
$607$ security changes collected from open-source software, from different
programming languages. The main contributions of the present work are:
\begin{itemize}
	\item An empirical study on the impact of security changes on software
	maintainability and what patterns need more attention.

@@

As an example, consider the refactoring of the Datagram Transport Layer Security
(DTLS) protocol implementation in OpenSSL\footnote{OpenSSL is a toolkit that
contains open-source implementations of the SSL and TLS cryptographic
protocols.} to address a vulnerability where an attacker could cause a Denial-of-Service
(DoS) attack by crafting DTLS handshake messages to trigger memory allocations
corresponding to large length values. This vulnerability is listed at the Common
Vulnerabilities and Exposures dictionary as CVE-$2014$-$3506$\footnote{CVE-$2014$-$3506$
details available at http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506
(Accessed on \today{})}. It is amongst the vulnerabilities studied in our
evaluation. The following snippet presents the changes performed on the
\emph{ssl/d1\_both.c} file\footnote{CVE-$2014$-$3506$ fix available  at
https://github.com/openssl/openssl/commit/\\1250f12613b61758675848f6600ebd914ccd7636
(Accessed on \today{})} by the OpenSSL developers to patch the vulnerability.

@@

\setcounter{lstannotation}{0}
\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
\\CVE-$2014$-$3506$ vulnerability},label={lst:vuln}]
+ static unsigned long dtls1_max_handshake_message_len(const SSL *s){ /*!\annotation{lst:func1}!*/
+  unsigned long max_len = DTLS1_HM_HEADER_LENGTH + SSL3_RT_MAX_ENCRYPTED_LENGTH;
+  if(max_len < (unsigned long)s->max_cert_list)
+     return s->max_cert_list;
+  return max_len;
+ }

@@

Every SSL/TLS connection begins with a handshake which is responsible by the negotiation between two parties.
DTLS provides a mechanism for fragmenting a handshake message over a number of packets, each of which can be
transmitted separately. In the example (Listing~\ref{lst:vuln}), all functions receive as input the
SSL session, \texttt{s}. In the \texttt{dtls1\_reassemble\_fragment} function the size of the fragment \texttt{msg\_hdr-$>$frag\_off+frag\_len} is checked against the handshake message size \texttt{msg\_hdr->msg\_len}
(line $13$) and then against the maximum handshake message size allowed by DTLS1 for \texttt{s} (line $16$-$21$).
However, it is not verified if \texttt{msg\_hdr->msg\_len} is lower than the allowed by DTLS. Thus, in line $37$,
when \texttt{msg\_len} is allocated to the fragment buffer, more memory than the allowed might be allocated. According to the message commit, $10$ handshake messages are allowed per DTLS connection. Hence, a remote attacker could leverage this issue to trigger
memory allocations corresponding to large length values and lead the system to memory exhaustion that could end
in a DoS attack. The code changes performed to fix the vulnerability were the following:

~\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was inserted to return the number of
maximum number of bytes allowed in a DTLS handshake message for \texttt{s}. The minimum is $16$KB but may be greater
if the maximum certificate list size requires it which is set by the user. More specifically, it checks if the
maximum certificate list size
(\texttt{s-$>$max\_cert\_list}) requires more length than the maximum length
(\texttt{max\_len}). If that condition holds, the function returns the size of
the certificate list, otherwise, it returns the maximum length. This function replaces lines $18$-$21$.

~\ref{lst:func5} and~\ref{lst:func4} add the check to different functions to ensure that the message
and the fragment length are never higher than the maximum allowed by DTSL. However, \ref{lst:func5} is responsible for patching the vulnerability.

@@

Often security flaws require refactoring code to make software more secure.
However, there is no evidence yet of how security refactorings impact the
maintainability of open-source software. Our suspicion is that developers tend
to introduce technical debt in their software when refactoring the codebase to
address a security flaw because they tend to choose the easiest path to solve
it. To address it, we compute the maintainability of $607$ commits using
\emph{Better Code Hub}. The same approach is applied to a randomly generated
dataset of regular commits --- baseline --- that we use to understand how
maintainability evolves when security refactorings are performed versus when
they are not.

@@

\end{lstlisting}
%
\section{Methodology}\label{sec:methodology}
%
The methodology used to measure the impact of security refactorings on the
maintainability of open-source software is discussed in this section. The methodology
comprehends the following steps, as illustrated in Figure~\ref{fig:met}:
\begin{enumerate}
	\item Extract relevant data from a dataset containing
$716$ security refactorings collected from open-source software available on
GitHub~\cite{Reis:2017:IJSSE}.
	\item A baseline of regular commits was
randomly collected from the list of projects of the main dataset to evaluate the impact of regular commits on the maintainability of open-source
software.
\item Use the Software Improvement Group's\footnote{SIG's website is available
at https://www.sig.eu/ (Accessed on \today{})} web-based source code analysis
service \emph{Better Code Hub} (BCH)\footnote{BCH's website is available at
https://bettercodehub.com/ (Accessed on \today{})} to quantify
maintainability for both security and regular commits.
\end{enumerate}

%
\begin{figure}[h]
 	\centering 	\includegraphics[width=0.5\textwidth]{figures/methodology.pdf}
 	\caption{Study Methodology}
	\label{fig:met}
\end{figure}
%
\subsection{Dataset}
%
To evaluate the impact of security refactorings on the maintainability of
open-source software, we use a dataset of security flaws which is the outcome of
mining $94$ GitHub projects. Reis and Abreu ($2017$) mined open-source
software aiming at the extraction of real examples --- created by real
developers --- to test and assess the performance of static analysis tools since
using hand-seeded test cases or mutations could lead to misleading assessments
of the capabilities of the tools~\cite{just2014mutants}. The study yielded a
dataset with $716$ test cases for $16$ security patterns. Some of the patterns
are based in the OWASP Top $10$ of $2013$~\cite{oswap:2013} and OWASP Top $10$ of
$2017$~\cite{oswap:2017}. Each test case of the
dataset is a triplet: the commit before the refactoring, the commit responsible
for the refactoring, and the snippets of code that differ from one version to
another (typically, called \textit{diffs}) --- where one can easily review the
code used to fix the security flaw. In this study, we focus on computing the
maintainability of the commits before and after the security refactoring to
evaluate if the impact was positive, negative or none.

@@

The $716$ refactorings in the dataset were run against the BCH toolset to
calculate their maintainability reports. However, there are refactorings
discarded from tour study due to limitations of BCH: in particular, lack of
language support and project size. The final dataset used in this paper contains
$607$ security refactorings labeled as one of the following categories (we
have decided to change the label to \textit{Miscellaneous} of the patterns in
the dataset containing less than $20$ instances):

\begin{itemize}
    \item \textbf{Using Components with Known Vulnerabilities.} The majority of
    software produced today integrates several components, such as libraries and
    frameworks, which may affect the software if developers use vulnerable
    versions - usually known and disclosed somewhere on the Internet.
    \textit{Based on the OWASP Top $10$ of $2017$.}

    \item \textbf{Broken Authentication \& Session Management.} Incorrectly implemented functionalities related to authentication and session management,
    allowing an attacker to gain access to session tokens, passwords, keys, and other sensitive data. \textit{Based on the OWASP Top $10$ of $2013$.}

    \item \textbf{Cross-Site Request Forgery.} Poor session tokens generation and management usually allow attackers to send forged HTTP requests including authentication information from the victim to the vulnerable web application.
    \textit{Based on the OWASP Top $10$ of $2013$.}

    \item \textbf{Denial-of-Service} Security flaws that can allow an attacker to flood or crashing services. This attacks usually occur when the system receives much traffic causing it to slow down and eventually stop. For
    example, flaws allowing attackers to trigger memory allocations corresponding
    to large length values (Listing~\ref{lst:vuln}).

    \item \textbf{Miscellaneous} This pattern comprises several other categories
    of security refactorings (e.g., \textit{path traversal and buffer overflow}).
    It contains several other security refactorings that do not have their own
    pattern yet. The patterns not satisfying the size requirement of more than $20$
    refactorings are contemplated here: \textit{Resource Leak},
    \textit{SHA-1 Hash Function}, \textit{Broken Access Control}, \textit{Sensitive Data Exposure},
    \textit{Insufficient Attack Protection}, \textit{Overflow}, \textit{Path Traversal}.

    \item \textbf{Cross-Site Scripting.} Lack of proper validation or escaping allow attackers to submit untrusted data to web browsers through malicious scripts that can hijack the user sessions or redirect the user to malicious sites. \textit{Based on the OWASP Top $10$ of $2017$.}

    \item \textbf{Injection.} When developers do not keep untrusted data separate from commands and queries. If an attacker sends a string that exploits the
    syntax of the interpreter, then an injection attack is possible (e.g., SQL and
    LDAP injection). \textit{Based on the OWASP Top $10$ of $2017$.}

@@

The dataset contemplates security flaws for more than $13$ different languages
being PHP ($39\%$) and C ($24\%$) the most prevalent ones; see Figure~\ref{fig:lang}.
To classify the applications in the dataset, we use a taxonomy for open-source
software from previous work~\cite{7816479}. Figure~\ref{fig:domain}
presents the projects domain distribution: \textit{Aplication Software} ($25$),
software that provides end-users with functional systems; \textit{Web libraries
and frameworks} ($20$); \textit{System Software} ($18$), software that provides
services and infrastructures (e.g., operating systems, servers and databases);
\textit{Software Tool} ($13$), software that supports development (e.g.,
programming languages, compilers, package managers, IDEs); and, \textit{Non-web
libraries and frameworks} ($18$), for desktop and mobile software.
%
Table~\ref{tab:dataset} presents the descriptive statistics of the $94$
open-source projects involved in this study, including the number of forks, the number of
stars, the number of watchers, the number of contributors, the number of commits, the number of
branches, the number of releases,  size (cf. GitHub it is the size of the whole
repository, including all of its history, in kilobytes), the number of issues, and the
number of pull requests.

%
\subsection{Security vs. Baseline Commits}
%
Previous studies attempted to measure the impact of regular refactorings on
open-source software maintainability~\cite{HEGEDUS2018313}. However,
there is no previous work focused on comparing the impact of security
refactorings with regular refactorings on maintainability.
We analyze the maintainability of regular commits
(i.e., commits not related to security fixes) and use them as a baseline.

The baseline dataset uses the security commits dataset as input. For each
security commit, one random commit is selected from the list of all commits of
that particular project. We originate the regular commits from the security
commits to ensure that differences in maintainability are not a consequence of
characteristics of different projects.
%
\subsection{Maintainability Analysis}

As said before, the web-based source code analysis service \emph{Better Code
Hub} (BCH) is used to collect the maintainability reports of the refactorings of
each project. Table~\ref{tab:guidelines} presents the $10$ guidelines proposed
by BCH's authors for delivering software that is not difficult to
maintain~\cite{Visser:2016:OREILLY}. During each guideline evaluation, the tool
determines the compliance towards one guideline by establishing limits for the
percentage of code allowed to be in each of the $4$ risk severity levels
(\emph{low risk}, \emph{medium risk}, \emph{high risk} and \emph{very high
risk}). If the project does not violate the thresholds, then it is compliant
with the guideline. These thresholds are calibrated by BCH using their own
data/experience --- open-source and closed software systems. If a project is
compliant with a guideline, it means that it is at least $65\%$ better than the
software used by BCH to calculate the thresholds\footnote{Check the answer to
\emph{How can I adjust the threshold for passing/not passing a guideline?} at
https://bettercodehub.com/docs/faq (Accessed on \today{})}.

Figure~\ref{fig:bchrep} shows an example of the report
provided by BCH for a project after finishing its evaluation. The example
refers to the OpenSSL CVE-$2014$-$3506$ vulnerability refactoring ---
as described as motivating example in Section~\ref{sec:motivation}. This
version of OpenSSL only complies with $1$ out of $10$ guidelines: \emph{Write
Clean Code}.

@@

\toprule
\textbf{10 Guidelines} & \textbf{Description}\\
\midrule
\textbf{Write Short Units of Code} & Limit code units to $15$ LOCs because smaller
 units are easier to understand, reuse and test them\\\midrule
\textbf{Write Simple Units of Code} & Limit branch points to $4$ per unit because
it makes units easier to test and modify \\\midrule
\textbf{Write Code Once} & Do not copy code because bugs tend to replicate at
multiple places (inefficient and error-prone)\\\midrule
\textbf{Keep Unit Interfaces Small} & Limit the number of parameters to at most
$4$ because it makes units easier to understand and reuse\\\midrule
\textbf{Separate Concerns in Modules} & Avoid large modules because changes in
loosely coupled databases are easier to oversee and execute\\\midrule
\textbf{Couple Architecture Components Loosely} & Minimize the amount of code
within modules that are exposed to modules in other components\\\midrule
\textbf{Keep Architecture Components Balanced} & Balancing the number of
components ease locating code and allow for isolated maintenance\\\midrule
\textbf{Keep your Codebase Small} & Reduce and avoid the system size because
small products are easier to manage and maintain\\\midrule
\textbf{Automate Tests} & Test your codebase because it makes development
predictable and less risky\\\midrule
\textbf{Write Clean Code} & Avoid producing software with code smells because
it is more likely to be maintainable in the future\\
\bottomrule
\end{tabular}
\label{tab:guidelines}
\end{table}

\begin{figure}[h]
 	\centering 	\includegraphics[width=0.5\textwidth]{figures/bch_report.png}
 	\caption{Maintainability report of OpenSSL's CVE-$2014$-$3506$ vulnerability
refactoring for the guideline \emph{Write Simple Units of Code} provided by
\emph{Better Code Hub}. This version of OpenSSL does not comply with the
guideline in the example since the bars do not reach the threshold points. This
example only complies with $\sfrac{1}{10}$ guidelines (\emph{Write Clean Code}).}
	\label{fig:bchrep}
\end{figure}
